\large{\bf Large effects and the infinitesimal model }

### 2023-12-03 review-responses.tex

\begin{quote}
Anyway, the results
in Sect 4 are of interest because they strongly constrain the possibilities to justify such
‘Gaussian-population approximation’ by any infinitesimal-like model (essentially to
mixtures of Cauchy distributions. The noted relation to blending inheritance is intriguing!).
My main concern is that especially this section needs are clearer description of the model,
and clear distinction between this ‘maximalist’ model (which has little theoretical
support) and the model investigated by Barton et al (2017). Also a much clearer summary
and discussion of the implications of the results shown here are needed. It should not be
assumed that readers have read Turelli (2017) and even less that they have studied
\citet{barton2017infinitesimal} in considerable detail!
\end{quote}

Thanks for the suggestions.
We also thought the section was useful (and also intruiging!),
which is why we left the section in.
(Also note it motivates the form of selection in our simulations.)
We share the reviewer's concern about context,
and have added some additional text
while trying not to get into the weeds too much
-- in particular, see the new text around \llname{clarification3}.

\begin{quote}
In Section 5, it is pointed out why the Barton et al. (2017) fails in the case of ‘Cauchy
noise’ (and presumably for any other limit law with $\alpha < 2$). In the absence of giving the
estimate in App D in Barton et al, the explanation around \llname{approx_eqn} is difficult to
follow. Please provide their estimate for easier comparison (do not assume that readers
studied Barton et al; assume that they scanned (parts) of it).
\end{quote}

\comment{We deliberately avoided giving the analogous estimate from Barton et al. (2017), as the expression is complicated and far from transparent (its definition spans several paragraphs). As such, we instead chose to focus on its salient features, namely that scales as $\frac{1}{\sqrt{M}}$ and grows (approximately) proportionally to the magnitude of the trait.  We have added a reference to the relevant appendix in Barton et al. (2017) for those sufficiently motivated, and have attempted to give some more intuition for the consequences of the difference in the error bounds.}

@@

\reply{Done \revref.
}

\begin{point}{}
Around \llname{cauchy_note}: It might be useful to mention earlier that the Cauchy distribution is an archetypical
example for an $\alpha$-stable distribution. The enthusiasm among biologists may be limited
though, because all the empirical trait distributions they find have mean and variance. I
don’t know if there are examples in which the variance grows (significantly) as the sample
size increases. In the last para of Sect 4.1 \revref, please specify ‘one will get convergence’. I
suppose this refers again to $M \to \infty$.
\end{point}

\reply{
    Good suggestion -- we've added the note about the Cauchy \llname{cauchy_note}.
    (We haven't tried to make biologists more enthusiastic, though,
    since we don't know of a situation in which strict blending inheritance seems terribly likely.)
    As for the note about ``one will get convergence'' -- it is as the number of generations increases,
    not as $M \to \infty$ (in fact, $M$ does not enter into this section at all);
    we've rewritten this statement to remind the reader
    about what's going on.
}

\begin{point}{\revref}
p 12, last display equation: Please relate this explicitly to the corresponding result in
Barton et al. (e.g., by stating their estimate).
\end{point}

\reply{
There isn't an analogous result in Barton et al., as they have assumed that all allele effects are bounded random variables rescaled by a factor of $\frac{1}{\sqrt{M}}$.  Thus, no matter how many alleles that one considers, none will have the large effects necessary to give information about that trait value.  We've added some text to clarify this difference at \revref.
}

### 2023-12-03 writeup.tex
TO DOs done, typos corrected.
Still need to read through and have improvements to make to 4 & 5.

To see what knowing the trait value $Z^{[j]}_{M}$ \revpoint{1}{16}
of a particular parent in the ancestral generation
tells us about the contribution of any given locus, $\eta^{[j]}_{\ell}$, to their trait value,
we follow the arguments in  \citet{Fisher1918} and \citet{barton2017infinitesimal} using Bayes' Theorem.
Proceeding informally, with the probabilities indicating probability density functions, we have 
\begin{align*}
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\} &= \frac{\P\{\eta^{[j]}_{\ell} = y; Z^{[j]} = z\}}{\P\{Z^{[j]} = z\}}\\
        &= \frac{\P\{Z^{[j]} = z\vert\eta^{[j]}_{\ell} = y\}}{\P\{Z^{[j]} = z\}}\P\{\eta^{[j]}_{\ell} = y\}\\
        &=  \frac{\P\{Z^{[j]} -\frac{1}{M} \eta^{[j]}_{\ell} = z - \frac{1}{M} y\}}{\P\{Z^{[j]} = z\}}\P\{\eta^{[j]}_{\ell} = y\}\\
\end{align*}
Now, $Z^{[j]} -\frac{1}{M} \eta^{[j]}_{\ell} = \frac{1}{M} \sum_{k = 1 : k \neq i}^{M} \eta^{[j]}_{k}$ is Cauchy-distributed with scale $\bar{\gamma} - \frac{\gamma_{\ell}}{M}$.  Using the probability density function for a Cauchy distributed random variable \eqref{eq:Cpdf}, a straightforward computation gives
\begin{align*}
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\}
    &=
    \frac{f\left(z - \frac{y}{M} ;0,\bar{\gamma}-\frac{\gamma_{\ell}}{M}\right)}{f(z;0,\bar{\gamma})} \P\{\eta^{[j]}_{\ell} = y\}\\
    &=
    \left(1
        +
        \frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})}
        \frac{1}{M}
        + \Oh\left(\frac{1}{M^{2}}\right)\right)
        \P\{\eta^{[j]}_{\ell} = y\}.
\end{align*}
Noting that $|(\bar{\gamma}^2 - z^2)/(\bar{\gamma}^2 + z^2)| \le 1$
and $|z/(\bar{\gamma}^2 + z^2)| \le 1/(2 \bar{\gamma})$
(since it is maximized at $z = \bar{\gamma}$),
so
\begin{align*}
        \left|
        \frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})}
        \right|
        \le \frac{|y|}{\gamma_\ell \bar{\gamma}} ,
\end{align*}
% Now, as $z \to 0$,
% $\frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})} \to \frac{\gamma_{\ell}}{\bar{\gamma}}$
% and as $z \to \pm \infty$,
% $\frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})} \to -\frac{\gamma_{\ell}}{\bar{\gamma}}$,
% \comment{I'm not sure where $\gamma_\ell$ went here}
% whereas $|\frac{\bar{\gamma}^{2}-z^{2}+2zy}{\bar{\gamma}^{2}+z^{2}}|$ 
% \[
% 	\frac{(\gamma_{\ell}^{2}+y^{2})\left(\sqrt{\gamma_{\ell}^{2}+y^{2}}-\gamma_{\ell}\right)}{\bar{\gamma}(\gamma_{\ell}^{2}+y^{2} - \gamma_{\ell}\sqrt{\gamma_{\ell}^{2}+y^{2}})},
% \]
% which is asymptotically equal to $\frac{|y|}{\bar{\gamma}}$ for large values of $|y|$, at
% \[
% 	z = -\frac{\bar{\gamma}}{y}(\gamma_{\ell} \pm \sqrt{\gamma_{\ell}^{2}+y^{2}}),  
% \]
% which is itself asymptotically equal to $\mp \sgn(y) \bar{\gamma}$ for large values of $|y|$.
we have that
\[
    \left|
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\} - \P\{\eta^{[j]}_{\ell} = y\}
    \right|
    \le \left(\left(1 + \frac{|y|}{\gamma_\ell \bar{\gamma}}\right) \frac{1}{M} + \Oh\left(\frac{1}{M^2}\right)\right)  \P\{\eta^{[j]}_{\ell} = y\}.
\]
% Unlike the case of bounded allelic effects, the bound is independent of the trait value $Z^{[j]}$. 
\llabel{approx_eqn}
In the case of bounded allelic effects, the error introduced by approximating the effect-size distribution conditional on the trait value by the unconditioned distribution is proportional to $|Z^{[j]}|$, so that extreme trait values are most informative of the size of allelic effects,
whereas here,
the bound is independent of the trait value,
and the error is maximized when the trait value is approximately equal to
the mean absolute deviation of its distribution, \ie the median value of $|Z^{[j]}|$.

The argument above shows for a typical locus, the effect of conditioning on the trait gives negligible information about the contribution from that site.   Recalling that
\[
    \P\left\{|\eta^{[j]}_{\ell}| > y\right\} =  1-\frac{2}{\pi} \arctan\left(\frac{y}{\gamma_{\ell}}\right) \sim \frac{2\gamma_{\ell}}{\pi y},
\]
we see that for any $\epsilon > 0$, and large values of $M$,
\[
    \P\left\{\frac{\eta^{[j]}_{\ell}}{\gamma_\ell \bar{\gamma} M} > \epsilon\right\} \sim \frac{2\gamma_{\ell}}{\pi \gamma_\ell \bar{\gamma} M \epsilon}.
\]
so the probability the error is larger than $\epsilon$ at any given locus decays rapidly as $M \to \infty$.

If, however, we consider not any one locus, but rather all loci, we get a different picture. 
The variables $\eta^{[j]}_{\ell}$ are independent, and thus so are the Bernoulli random variables 
\begin{equation}
	\1^{M}_{\ell} = \begin{cases}
        1 & \text{if $\frac{\eta^{[j]}_{\ell}}{\bar{\gamma} M} > \epsilon$, and}\\
		0 & \text{otherwise.}
	\end{cases}
\end{equation}
Now, since
\[
	\lim_{M \to \infty} \sum_{\ell = 1}^{M} \frac{2\gamma_{\ell}}{\pi \bar{\gamma} M \epsilon} = \frac{2}{\pi \epsilon},
\]
the number of loci at which the magnitude of the effect size exceeds $M \epsilon$, which is 
$\sum_{\ell = 1}^{M} \1^{M}_{\ell}$,
converges weakly as $M \to \infty$ to a Poisson random variable with mean $2/(\pi \epsilon)$ (see \eg Theorem 2.6.1 in \cite{Durrett2005}).  In particular, for large values of $M$ and fixed $\epsilon > 0$,\revpoint{1}{17}
\begin{align*}
    \P\left\{\max_{\ell=1}^{M} \frac{\eta^{[j]}_{\ell}}{\bar{\gamma} M} > \epsilon\right\}  \sim 1-e^{-\frac{2}{\pi \bar{\gamma} \epsilon}},
\end{align*}
which approaches 1 as $\epsilon \to 0$, while the expected number of $M$ alleles such that $\frac{\eta^{[j]}_{\ell}}{\bar{\gamma} M} > \epsilon$ is asymptotically $\frac{2}{\pi\epsilon}$.

>>>>

To see what knowing the trait value $Z^{[j]}_{M}$ \revpoint{1}{16}
of a particular parent in the ancestral generation
tells us about the contribution of any given locus, $\eta^{[j]}_{\ell}$, to their trait value,
we follow the arguments in  \citet{Fisher1918} and \citet{barton2017infinitesimal} using Bayes' Theorem.
Proceeding informally, with the probabilities indicating probability density functions, we have 
\begin{align*}
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\} &= \frac{\P\{\eta^{[j]}_{\ell} = y; Z^{[j]} = z\}}{\P\{Z^{[j]} = z\}}\\
        &= \frac{\P\{Z^{[j]} = z\vert\eta^{[j]}_{\ell} = y\}}{\P\{Z^{[j]} = z\}}\P\{\eta^{[j]}_{\ell} = y\}\\
        &=  \frac{\P\{Z^{[j]} -\frac{1}{M} \eta^{[j]}_{\ell} = z - \frac{1}{M} y\}}{\P\{Z^{[j]} = z\}}\P\{\eta^{[j]}_{\ell} = y\}\\
\end{align*}
Now, $Z^{[j]} -\frac{1}{M} \eta^{[j]}_{\ell} = \frac{1}{M} \sum_{k = 1 : k \neq i}^{M} \eta^{[j]}_{k}$ is Cauchy-distributed with scale $\bar{\gamma} - \frac{\gamma_{\ell}}{M}$.  Using the probability density function for a Cauchy distributed random variable \eqref{eq:Cpdf}, a straightforward computation gives
\begin{align*}
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\}
    &=
    \frac{f\left(z - \frac{y}{M} ;0,\bar{\gamma}-\frac{\gamma_{\ell}}{M}\right)}{f(z;0,\bar{\gamma})} \P\{\eta^{[j]}_{\ell} = y\}\\
    &=
    \left(1
        +
        \frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2\bar{\gamma} zy}{\bar{\gamma}(\bar{\gamma}^{2}+z^{2})}
        \frac{1}{M}
        + \Oh\left(\frac{1}{M^{2}}\right)\right)
        \P\{\eta^{[j]}_{\ell} = y\}.
\end{align*}
Noting that $|(\bar{\gamma}^2 - z^2)/(\bar{\gamma}^2 + z^2)| \le 1$
and $|z/(\bar{\gamma}^2 + z^2)| \le 1/(2 \bar{\gamma})$
(since it is maximized at $z = \bar{\gamma}$),
so
\begin{align*}
        \left|
        \frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2\bar{\gamma}zy}{\bar{\gamma}(\bar{\gamma}^{2}+z^{2})}
        \right|
        \le \frac{\gamma_{\ell}}{\bar{\gamma}} + \frac{|y|}{\bar{\gamma}},
\end{align*}
% Now, as $z \to 0$,
% $\frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})} \to \frac{\gamma_{\ell}}{\bar{\gamma}}$
% and as $z \to \pm \infty$,
% $\frac{\gamma_{\ell}(\bar{\gamma}^{2}-z^{2})+2zy}{\gamma_{\ell}(\bar{\gamma}^{2}+z^{2})} \to -\frac{\gamma_{\ell}}{\bar{\gamma}}$,
% \comment{I'm not sure where $\gamma_\ell$ went here}
% whereas $|\frac{\bar{\gamma}^{2}-z^{2}+2zy}{\bar{\gamma}^{2}+z^{2}}|$ 
% \[
% 	\frac{(\gamma_{\ell}^{2}+y^{2})\left(\sqrt{\gamma_{\ell}^{2}+y^{2}}-\gamma_{\ell}\right)}{\bar{\gamma}(\gamma_{\ell}^{2}+y^{2} - \gamma_{\ell}\sqrt{\gamma_{\ell}^{2}+y^{2}})},
% \]
% which is asymptotically equal to $\frac{|y|}{\bar{\gamma}}$ for large values of $|y|$, at
% \[
% 	z = -\frac{\bar{\gamma}}{y}(\gamma_{\ell} \pm \sqrt{\gamma_{\ell}^{2}+y^{2}}),  
% \]
% which is itself asymptotically equal to $\mp \sgn(y) \bar{\gamma}$ for large values of $|y|$.
we have that
\[
    \left|
    \P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\} - \P\{\eta^{[j]}_{\ell} = y\}
    \right|
    \le \left(\left(\frac{\gamma_{\ell}}{\bar{\gamma}} + \frac{|y|}{\bar{\gamma}}\right) \frac{1}{M} + \Oh\left(\frac{1}{M^2}\right)\right)  \P\{\eta^{[j]}_{\ell} = y\}.
\]
% Unlike the case of bounded allelic effects, the bound is independent of the trait value $Z^{[j]}$. 
\llabel{approx_eqn}

When allelic effects have magnitude bounded by  $\frac{B}{\sqrt{M}}$, (see Appendix C in \citet{barton2017infinitesimal}), the error introduced by approximating the effect-size distribution conditional on the trait value by the unconditioned distribution takes the form
\[
	\left|\P\{\eta^{[j]}_{\ell} = y \vert Z^{[j]} = z\} - \P\{\eta^{[j]}_{\ell} = y\} \right| \leq \frac{1}{\sqrt{M}} C(\hat{\sigma}_{M}^{2},z) \P\{\eta^{[j]}_{\ell} = y\},
\]
where $\hat{\sigma}_{M}$ is the variance in the trait value $Z^{[j]}$ and $C(\hat{\sigma}_{M}^{2},z)$ is a somewhat complicated expression that is approximately proportional to $z$, but independent of the allelic effect size $y$ (Equation (12), \textit{loc. cit.}).  Thus, the trait value $Z^{[j]}$ is most informative of the size of the allelic effects at any given locus when it differs significantly from its mean: intuitively, when allele effects are uniformly bounded, a large trait value is evidence for many atypically large allelic effects.  Interestingly, for the Cauchy distribution, we have the opposite: the error bound is independent of the trait value, $Z^{[j]} = z$, but is proportional to the allele effect size, $\eta^{[j]}_{\ell} = y$, at the locus under consideration (indeed, the error is maximized when the trait value is approximately equal to the mean absolute deviation of its distribution, \ie the median value of $|Z^{[j]}|$, rather than at extreme trait values).  This suggests that with Cauchy-distributed allelic effects, knowing the trait value gives evidence for large allelic effects at individual loci.

To elaborate, the argument above shows for a typical locus, the effect of conditioning on the trait gives negligible information about the contribution from that site.   Recalling that
\[
    \P\left\{|\eta^{[j]}_{\ell}| > y\right\} =  1-\frac{2}{\pi} \arctan\left(\frac{y}{\gamma_{\ell}}\right) \sim \frac{2\gamma_{\ell}}{\pi y} \quad \text{for large values of $y$},
\]
we see that for any $\epsilon > 0$, and large values of $M$,
\[
    \P\left\{\frac{|\eta^{[j]}_{\ell}|}{\bar{\gamma} M} > \epsilon\right\} \sim \frac{2\gamma_{\ell}}{\pi \bar{\gamma} M \epsilon}.
\]
so the probability the error is larger than $\epsilon$ at any given locus decays rapidly as $M \to \infty$.

If, however, we consider not any one locus, but rather all loci, we get a different picture. 
The variables $\eta^{[j]}_{\ell}$ are independent, and thus so are the Bernoulli random variables 
\begin{equation}
	\1^{M}_{\ell} = \begin{cases}
        1 & \text{if $\frac{|\eta^{[j]}_{\ell}|}{\bar{\gamma} M} > \epsilon$, and}\\
		0 & \text{otherwise.}
	\end{cases}
\end{equation}
Now, since
\[
	\lim_{M \to \infty} \sum_{\ell = 1}^{M} \frac{2 \gamma_{\ell}}{\pi \bar{\gamma} M \epsilon} = \frac{2}{\pi \epsilon},
\]
the number of loci at which the magnitude of the scaled effect size exceeds $\bar{\gamma} \epsilon$, which is 
$\sum_{\ell = 1}^{M} \1^{M}_{\ell}$,
converges weakly as $M \to \infty$ to a Poisson random variable with mean $2/(\pi \epsilon)$ (see \eg Theorem 2.6.1 in \cite{Durrett2005}).  In particular, for large values of $M$ and fixed $\epsilon > 0$,\revpoint{1}{17}
\begin{align*}
    \P\left\{\max_{\ell=1}^{M} \frac{|\eta^{[j]}_{\ell}|}{\bar{\gamma} M} > \epsilon\right\}  \sim 1-e^{-\frac{2}{\pi \epsilon}},
\end{align*}
which approaches 1 as $\epsilon \to 0$, while the expected number of $M$ alleles such that $\frac{\eta^{[j]}_{\ell}}{\bar{\gamma} M} > \epsilon$ is asymptotically $\frac{2}{\pi\epsilon}$. 
By contrast, in \citet{Fisher1918} and \citet{barton2017infinitesimal} \revpoint{2}{10}, the scaled allelic effect at any locus $\ell$ has magnitude bounded by $\frac{B}{\sqrt{M}}$, and thus, no matter how many alleles one considers, there will be no large effect alleles that are informative of the trait value.
